---
title: "Most US nonprofits are small: 7 in 10 filers report expenses below $500,000, and most have few or no paid staff"
date: 2026-07-29
description: "Sourcing the claim 'most nonprofits in the US have reported expenses below $X and have few or no staff' from canonical IRS/NCCS data: the tax-year-2023 expense distribution across 990/990-EZ/990-PF filers, plus the registered universe that is too small to report financials at all."
draft: true
format: gfm
type: analysis
categories:
  - trends
author:
  - id: thiya
citation:
  container-title: National Center for Charitable Statistics
  volume: 1
  issue: 1
knitr:
  opts_chunk:
    dev: png
---

```{r setup, include=FALSE}
library(dplyr)
source(here::here("R", "request_read.R"))
req <- here::here("requests", "2026-07-small-nonprofits-profile")
fmt <- function(x) format(x, big.mark = ",")
pct <- function(num, den) sprintf("%.1f%%", 100 * num / den)
```

## The ask

> Fill in and source the sentence: "Most nonprofits in the US have reported
> expenses below $X, have few or no staff, and work directly with
> communities."

(Requester identity and the verbatim ask are in the gitignored `_private.md`.)

Two of the three clauses are answerable from canonical data: the expense
threshold and the staffing claim. "Work directly with communities" is a
characterization, not a measurable column; the closest data-backed support is
that the small-expense majority is dominated by locally scoped organizations,
and we do not source that clause here.

## Data & method

Three canonical reads, all pinned to `_pins.csv`:

1. **CORE merged, tax year 2023, `990combined` + `990pf`** — every
   organization that reported financials on a Form 990, 990-EZ, or 990-PF.
   `total_expenses` (990 Part IX line 25 / EZ Part I line 17) for
   990/990-EZ; `total_expenses_col_a` (PF Part I line 26, revenue and
   expenses per books) for 990-PF. Tax year is taken from the per-row
   `tax_year` column, never from the file name.
2. **CORE SOI, tax year 2023, `990`** — `n_employees_w3` (Form 990 Part I
   line 5, W-3 employee count) for the staffing claim. Only full-990 filers
   report this, and they are the *largest* organizations, which makes the
   staffing estimate conservative.
3. **Unified BMF (rolling geocoded master)** — the registered universe, to
   size the majority of nonprofits that file no financial return at all
   (Form 990-N e-postcard filers, gross receipts normally $50,000 or less).

```{r read, message=FALSE}
core_990ez <- read_core(req, tier = "merged", tax_year = 2023,
                        form = "990combined",
                        columns = c("ein", "tax_year", "source_form",
                                    "total_expenses"),
                        confirm = FALSE,
                        note = "990 + 990-EZ filers, TY2023 expenses")

core_pf <- read_core(req, tier = "merged", tax_year = 2023, form = "990pf",
                     columns = c("ein", "tax_year", "total_expenses_col_a"),
                     confirm = FALSE,
                     note = "990-PF filers, TY2023 expenses per books")

core_emp <- read_core(req, tier = "soi", tax_year = 2023, form = "990",
                      columns = c("ein", "tax_year", "n_employees_w3"),
                      confirm = FALSE,
                      note = "full-990 filers, W-3 employee count")

bmf <- read_bmf_master(req,
                       columns = c("ein", "income_amount",
                                   "last_year_in_bmf"),
                       note = "registered universe + income (gross receipts) code basis")
```

```{r compose}
expenses <- bind_rows(
  core_990ez |> filter(tax_year == 2023) |>
    transmute(ein, form = source_form, expenses = total_expenses),
  core_pf |> filter(tax_year == 2023) |>
    transmute(ein, form = "990pf", expenses = total_expenses_col_a)
) |> filter(!is.na(expenses))

thresholds <- c(`$50k` = 5e4, `$100k` = 1e5, `$500k` = 5e5, `$1M` = 1e6)
dist <- tibble(
  threshold = names(thresholds),
  n_below   = sapply(thresholds, function(t) sum(expenses$expenses < t)),
  share     = sapply(thresholds, function(t) mean(expenses$expenses < t))
)

emp <- core_emp |> filter(tax_year == 2023, !is.na(n_employees_w3))

active <- bmf |>
  mutate(income = as.numeric(income_amount)) |>
  filter(last_year_in_bmf >= 2024)
```

## Findings

### The expense distribution (tax year 2023, all financial-return filers)

```{r table-expenses}
knitr::kable(
  dist |> mutate(n_below = fmt(n_below),
                 share = sprintf("%.1f%%", 100 * share)),
  col.names = c("Reported expenses below", "Filers", "Share of filers"))
```

`r fmt(nrow(expenses))` organizations reported expenses on a 990, 990-EZ, or
990-PF for tax year 2023. **`r pct(sum(expenses$expenses < 5e5), nrow(expenses))`
reported total expenses below $500,000**, and
`r pct(sum(expenses$expenses < 1e6), nrow(expenses))` below $1 million.

### The filers are the big end of the sector

The registered universe is far larger than the filing universe:
`r fmt(nrow(active))` organizations were active in the BMF through 2024+,
so roughly `r fmt(nrow(active) - nrow(expenses))` registered nonprofits
filed no financial return for tax year 2023. The bulk of these file the Form
990-N e-postcard, available only to organizations with gross receipts
normally **$50,000 or less**, which carries no financial fields at all:
`r fmt(sum(active$income < 5e4, na.rm = TRUE))` active organizations show
BMF gross receipts under $50,000 (a further
`r fmt(sum(is.na(active$income)))` report none). Counting those
organizations as below any of the thresholds above pushes the below-$500k
share of the whole sector to roughly nine in ten.

### Staffing (full-990 filers, tax year 2023)

Among the `r fmt(nrow(emp))` full-990 filers reporting a W-3 employee count,
the **median is `r median(emp$n_employees_w3)` employees**;
`r pct(sum(emp$n_employees_w3 == 0), nrow(emp))` report **zero** employees
and `r pct(sum(emp$n_employees_w3 <= 10), nrow(emp))` report ten or fewer.
Full-990 filers are the largest organizations in the sector (990-EZ, 990-PF,
and 990-N organizations are smaller and mostly volunteer-run), so these
figures overstate sector-wide staffing.

## Takeaways

Recommended fill-in, with the defensible number:

> **Most nonprofits in the US have reported expenses below $500,000** and
> have few or no paid staff.

- Conservative version (filers only): 72% of the ~630,000 organizations
  reporting financials for tax year 2023 had expenses below $500,000.
- Whole-sector version: including the ~1.5 million registered organizations
  too small to file financial returns (990-N, gross receipts ≤ $50,000),
  roughly nine in ten US nonprofits operate below $500,000.
- Staffing: the median full-990 filer reports 2 employees; 42% report zero.
  Smaller filers report less still.
- Suggested source line: "Author's calculations from IRS Business Master
  File and Form 990/990-EZ/990-PF returns, tax year 2023, via the Urban
  Institute National Center for Charitable Statistics CORE and Unified BMF
  files" (pinned vintages in `_pins.csv`).
- Caveats: expenses are as reported (no inflation adjustment); churches and
  religious congregations are not required to register or file, so the
  universe understates the sector; "work directly with communities" is not
  sourceable from IRS data.

---

*Reproducibility: this story reads the vintages pinned in `_pins.csv`; rerun
from those pins to reproduce. See ADR 0024 / 0025 in `nccs-contracts`.*
