---
title: "Nonprofits in the Milwaukee Metro Area"
subtitle: "A guide to the attached data — counts, focus areas, and revenue, 1989–2023"
date: 2026-06-05
format:
  html:
    theme: cosmo
    toc: true
    toc-title: "On this page"
    embed-resources: true
    fig-width: 8
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
library(knitr)
D <- function(f) read.csv(file.path("data", f), check.names = FALSE, stringsAsFactors = FALSE)
comma <- function(x) formatC(x, format = "d", big.mark = ",")
dollarsM <- function(x) paste0("$", formatC(x, format = "d", big.mark = ","), "M")
```

## What this is

This package answers five questions about nonprofit organizations in the
**Milwaukee metro area** — both today and back to 1989:

1. How many nonprofits there are, overall and by county
2. How many there are by **focus area** (what they do)
3. Their total **revenue by focus area**
4. Their total **revenue by source** (where the money comes from)
5. Revenue broken down by focus area **and** source together

**The metro area** is the four-county Milwaukee–Waukesha Metropolitan Statistical
Area as defined by the U.S. Office of Management and Budget: **Milwaukee,
Waukesha, Ozaukee, and Washington** counties.

**The data** comes entirely from public IRS records, organized by the National
Center for Charitable Statistics (NCCS): the IRS Business Master File (the list
of all registered nonprofits) for the counts, and nonprofits' annual Form 990
filings for the revenue figures. Nothing here is confidential.

## The headline numbers

```{r headline}
yr  <- D("nonprofits_by_year_1989-2026.csv")
cty <- D("nonprofits_by_county_2026.csv")
n_now <- yr$n_nonprofits[yr$year == 2026]
n_89  <- yr$n_nonprofits[yr$year == 1989]
```

- **About `r comma(n_now)` registered nonprofits** operate in the metro today —
  roughly **six times** the `r comma(n_89)` on record in 1989.
- **Health care organizations dominate the sector's revenue.** In 2023, hospitals
  and health systems alone reported about **$10 billion** of the roughly **$27
  billion** in total public-charity revenue across the metro.
- **Most revenue is earned, not donated.** Program service revenue (fees for
  services — hospital care, tuition, and the like) is about **80%** of the total;
  contributions and grants are the next-largest source.
- **Private foundations** in the metro add roughly **$1.2 billion** more in
  annual revenue (2023), reported separately because they file a different tax
  form.

## How many nonprofits, and where

### By county (current)

```{r county}
k <- cty; k$n_nonprofits <- comma(k$n_nonprofits)
kable(k, col.names = c("County", "Nonprofits"), align = c("l","r"))
```

### Over time (registered nonprofits in the metro)

```{r year}
sel <- yr[yr$year %in% c(1989,1995,2000,2005,2010,2015,2020,2023,2026), ]
sel$n_nonprofits <- comma(sel$n_nonprofits)
kable(sel, col.names = c("Year", "Nonprofits"), align = c("l","r"))
```

*See `nonprofits_by_year_1989-2026.csv` for every year. Counts for the 1990s and
2000s undercount somewhat — see "How to read these numbers" below.*

## What they do — focus areas

```{r focus}
fa <- D("nonprofits_by_focus_area_2026.csv")
fa <- fa[order(-fa$n_nonprofits), ]; fa$n_nonprofits <- comma(fa$n_nonprofits)
kable(fa, col.names = c("Focus area", "Nonprofits"), align = c("l","r"))
```

## Revenue by focus area (2023)

```{r revfocus}
rf <- D("revenue_by_focus_area_2023.csv")
rf$total_revenue_usd_millions <- dollarsM(rf$total_revenue_usd_millions)
rf$n_filers <- comma(rf$n_filers)
kable(rf, col.names = c("Focus area", "Filers", "Total revenue"), align = c("l","r","r"))
```

## Revenue by source, over time

```{r revsource}
rs <- D("revenue_by_source_by_year_1989-2023.csv")
sel <- rs[rs$year %in% c(1989,2000,2010,2019,2023), ]
for (c in setdiff(names(sel),"year")) sel[[c]] <- dollarsM(sel[[c]])
kable(sel, col.names = c("Year","Contributions & grants","Program services",
      "Fundraising events (net)","Investment","Other","Total"), align = "r")
```

*Dollar figures in millions. "Program services" = fees earned for an
organization's mission work (e.g. hospital and clinic charges, tuition). Full
year-by-year detail is in `revenue_by_source_by_year_1989-2023.csv`.*

## How to read these numbers

- **Two different groups are being counted.** The *counts* cover every registered
  nonprofit (all types). The *revenue* figures cover only organizations that file
  a financial tax return — **Form 990 or 990-EZ** for public charities, and
  **Form 990-PF** for private foundations (reported separately). The smallest
  nonprofits file a postcard return with no financials, so they appear in the
  counts but not the revenue totals.
- **Older years undercount, especially by county.** Placing an organization in a
  county requires a usable address on file. Coverage is essentially complete for
  recent years but lower for the distant past (about 40% in 1989, rising to ~100%
  today). Read the 1990s/2000s figures as **trends and minimums**, not exact
  totals. The file `data_coverage_by_year.csv` documents this.
- **Dollars are shown two ways.** "Nominal" is dollars as reported each year;
  "real 2024" adjusts for inflation so years can be compared on equal footing.
- **An "Unknown, Unclassified" focus area** appears because the IRS classification
  is missing for some organizations (including a few very large ones). It is shown
  as its own line rather than hidden.

## Files included

```{r files}
files <- data.frame(
  File = c(
    "nonprofits_by_year_1989-2026.csv",
    "nonprofits_by_county_2026.csv",
    "nonprofits_by_focus_area_2026.csv",
    "revenue_by_focus_area_2023.csv",
    "revenue_by_source_by_year_1989-2023.csv",
    "revenue_by_focus_and_source_2023.csv",
    "total_revenue_public_charities_by_year_1989-2023.csv",
    "total_revenue_private_foundations_by_year_1989-2023.csv",
    "data_coverage_by_year.csv"),
  Contents = c(
    "Number of registered nonprofits in the metro, each year 1989–2026.",
    "Number of registered nonprofits today, by county.",
    "Number of registered nonprofits today, by focus area.",
    "2023 revenue by focus area (Form 990/990-EZ public charities).",
    "Revenue split by source, every year 1989–2023 (public charities).",
    "2023 revenue by focus area AND source together (public charities).",
    "Total public-charity (990/990-EZ) revenue by year, nominal and inflation-adjusted.",
    "Total private-foundation (990-PF) revenue by year, nominal and inflation-adjusted.",
    "Diagnostic: share of nonprofits with a usable county address, by year (the coverage caveat above)."),
  stringsAsFactors = FALSE)
kable(files, align = c("l","l"))
```

**Common columns:** `focus_area` uses the NCCS NTEEv2 subsector names (Health,
Education, Human Services, Arts, etc.); `n_nonprofits` / `n_filers` are counts;
columns named for a revenue source or ending in `usd_millions` are dollar amounts
**in millions**.

---

*Prepared by the National Center for Charitable Statistics. Source: IRS Business
Master File and Form 990 series via the `nccsdata` package; geography via the NCCS
county and metro-area crosswalks (2023 vintage). Reproducible from the pinned data
versions recorded with the analysis.*
</content>
