---
# --- Story front-matter (mirrors nccs/_stories so promotion is a clean move).
# Keep `draft: true` until the request clears the three gates in checklist.md.
title: "TITLE — what the request reveals, not who asked"
date: 2026-06-05
description: "One or two sentences a non-specialist would read."
draft: true
format: gfm
type: analysis          # analysis | methods
categories:
  - geography
author:
  - id: thiya
# Required before promotion: cite the pinned vintages this analysis read.
# Fill from _pins.csv (the `vintage` column) once your reads have run.
citation:
  container-title: National Center for Charitable Statistics
  volume: 1
  issue: 1
---

```{r setup, include=FALSE}
library(dplyr)
source(here::here("R", "request_read.R"))

# This request's folder (used to write _pins.csv). Update the slug.
req <- here::here("requests", "2026-06-template")
```

## The ask

> Restate the request in one or two lines — the *question*, not the requester.
> (Requester identity / any confidential specifics stay out of the `.qmd` —
> see `checklist.md`.)

## Data

Read canonical artifacts through the helpers — every read is pinned to
`_pins.csv`. Compose geography by **joining the published crosswalks**; do not
re-derive county / CBSA identity here.

```{r read}
# BMF (rolling geocoded master) — or read_bmf_vintage(req, "2023_07") to pin a snapshot
bmf <- read_bmf_master(req, state = "CT")

# Geography: label join for most states; coordinate join for Connecticut.
xwalk_county <- read_crosswalk(req, "county-fips")
# xwalk_cbsa   <- read_crosswalk(req, "cbsa")
# xwalk_ct     <- read_crosswalk(req, "ct-planning-region")   # CT only, by lat/lon

# core / efile as the request needs:
# core <- read_core(req, tier = "merged")
```

```{r compose}
# Join geography onto the raw labels / coordinates. Example (label grain):
# resolved <- bmf |>
#   left_join(xwalk_county, by = c("geo_state_abbr", "geo_county"))
#
# Connecticut: round (geo_lat, geo_lon) to 0.01 deg and join xwalk_ct
# (its retired county labels are deferred in county-fips — see ADR 0023).
```

## Findings

Write the analysis. Tables and figures render to SVG (per `_quarto.yml`).

## Takeaways

- Lead with the substantive finding.
- Note any coverage caveats (e.g. ambiguous/unresolved geo labels).

---

*Reproducibility: this story reads the vintages pinned in `_pins.csv`; rerun
from those pins to reproduce. See ADR 0024 / 0025 in `nccs-contracts`.*
